Relational Parametricity for Algebraically Indexed Types

### 2012-07-03 paper/section-general-framework.tex
Additions to the general framework, and fill in some of the examples
- Add a subsection on the construction of relational environments (unfinished)
- Fill in the monoid indexed types section
- Rearrange the example sections a bit

\begin{example}[Two-Dimensional Geometry]
  The two-dimensional geometry system has a sort for each of the
  geometric groups that we mentioned in
  \autoref{sec:motivating-examples}. Thus, $\SortSet =
  \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$. For the
  index operations, we have additive group structure on
  $\SynTransl{2}$, multiplicative group structure on $\SynGL{1}$,
  $\SynGL{2}$ and $\SynOrth{2}$, injections from $\SynOrth{2}$ and
  $\SynGL{1}$ into $\SynGL{2}$, determinant, and absolute value. Thus,
  $\IndexOpSet = \{ 0, +, -, 1_G, -\cdot_G-, -^{-1_G}, \iota_O,
  \iota_1, \det, |\cdot| \}$, where $G \in \{\SynGL{1}, \SynGL{2},
  \SynOrth{2}\}$, with the following assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl{2}) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
      \indexOpArity(|\cdot|) & ([\SynGL{1}], \SynGL{1})
    \end{array}
  \end{displaymath}
  The intended interpretations of the top three pairs of operations
  are group unit, group combination and group negation,
  respectively. When we introduce equational theories for index
  expressions in \autoref{sec:type-equality} below, we will impose the
  (abelian) group laws. For this example, we have $\PrimTypeSet = \{
  \tyPrimNm{vec}, \tyPrimNm{real} \}$, with
  $\primTyArity(\tyPrimNm{vec}) = [\SynGL{2}, \SynTransl{2}]$ and
  $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
\end{example}

% \begin{example}[Logical Information Flow]
%   \fixme{Do this}
% \end{example}

% \begin{example}[Units of Measure]
%   To instantiate our general framework as Kennedy's original units
%   of measure system, we would take the following settings. For the
%   collection of sorts we take $\SortSet = \{ \mathsf{unit} \}$. For
%   the collection of index operations, we take $\mathit{IndexOp} = \{1,
%   -\cdot-, -^{-1}\}$, with:
%   \begin{eqnarray*}
%     \indexOpArity{(1)} &=& ([],\mathsf{unit}) \\
%     \indexOpArity{(-\cdot-)} & = & ([\mathsf{unit},\mathsf{unit}],\mathsf{unit}) \\
%     \indexOpArity{(-^{-1})} & = & ([\mathsf{unit}],\mathsf{unit})
%   \end{eqnarray*}
%   The intended interpretation of these operations is group unit, group
%   multiplication and group inverse, respectively. When we discuss
%   equational theories for index expressions in
%   \autoref{sec:type-equality} below, we will impose the abelian group
%   laws on these operations.

%   The primitive types for the units of measure example are taken to be
%   $\PrimTypeSet = \{ \mathsf{num} \}$, with
%   $\primTyArity(\mathsf{num}) = [\mathsf{unit}]$.
% \end{example}

@@

  \textbf{Well-sorted index expressions}
  \begin{mathpar}
    \inferrule* [right=IVar]
    {i : s \in \Delta}
    {\Delta \vdash i : s}

@@

    \inferrule* [right=TyExist]
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \exists i \mathord: s. A \isType}
  \end{mathpar}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple: either an index expression
is a variable that appears in the context (rule \TirName{IVar}), or it
is an application of an index operation taken from $\IndexOpSet$ to
other index expressions (rule \TirName{IOp}). The rules for
well-indexed types include the usual rules for constructing types of
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index
expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} permits the
formation of universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types are constructed using the \TirName{TyExist}
rule, and allow for abstraction by hiding. %FIXME: think about the introduction of existentials more, perhaps forward ref to their use

@@

\begin{example}[Two-Dimensional Geometry]
  \fixme{Tie this example back to section 2}
  Translations form an abelian group under addition, so we assume the
  abelian group axioms for translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of an orthgonal transformation is always $1$, so we also
  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

% \begin{example}[Logical Information Flow]
%   \fixme{Do this}
% \end{example}

% \begin{example}[Units of Measure]
%   For the units of measure system, seen as an instantiation of our
%   general framework, the set $\IndexAxiomSet$ consists of the abelian
%   group axioms:
%   \begin{displaymath}
%     \begin{array}{l}
%       u : \mathsf{unit} \vdash u \cdot 1 \stackrel{ax}\equiv u : \mathsf{unit} \\
%       u_1, u_2, u_3 : \mathsf{unit} \vdash u_1 \cdot (u_2 \cdot u_3) \stackrel{ax}\equiv (u_1 \cdot u_2) \cdot u_3 : \mathsf{unit} \\
%       u_1, u_2 : \mathsf{unit} \vdash u_1 \cdot u_2 \stackrel{ax}\equiv u_2 \cdot u_1 : \mathsf{unit} \\
%       u : \mathsf{unit} \vdash u \cdot u^{-1} \stackrel{ax}\equiv 1 : \mathsf{unit} \\
%     \end{array}
%   \end{displaymath}
% \end{example}

@@

The defining feature of the index erasure interpretation is that the
semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index expressions $e_1,...,e_n$. Accordingly, we assume that each
primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$. We extend this assignment to every
well-indexed type by induction on the type structure:
\begin{displaymath}
  \begin{array}{@{\hspace{0em}}c@{\hspace{0em}}c}
    \begin{array}{@{}l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A} \\
      \tySem{\exists i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
The interpretation of the unit type is a chosen one element set, and
the interpretations of the function, tuple and sum types are simply
the corresponding constructions on sets. The interpretations of the
universal and existential quantifiers ignore the indexing: the
interpretation of the type $\forall i\mathord:s.A$ is exactly the
interpretation of the type $A$, and likewise for $\exists
i\mathord:s.A$.

@@

% \begin{example}[Logical Information Flow]
%   \fixme{do this}
% \end{example}

@@

For the unit, tuple, sum and function types we will define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types require an interpretation of index
contexts. We assign interpretations to index contexts in terms of sets
of \emph{relational environments}, which we now define.

A relation environment for a context $\Delta$ is a (dependently typed)
function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
instantiation of its arguments $\sigma$, assigns a binary relation
$\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure interpretation of
$\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
Every relation environment must respect index expression equality: for
any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
substitutions $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.

% FIXME: an example?

@@

% FIXME: fix the notation here for interpretation of index contexts:
% should use some sort of semantic brackets

As we stated above, index contexts $\Delta$ are interpreted as certain
sets of relation environments, i.e.,~subsets of
$\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
we assume for our system, we may have different sets of relational
environments that enforce different invariants. For the general
framework, we assume that, for each index context $\Delta$, we are
given a set of relation environments $\relEnv{E}(\Delta)$. Note that
this assignment of sets of relation environments to index contexts is
not assumed to be compositional: it is not necessarily the case that
the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
defined in terms of $\relEnv{E}(\Delta_1)$ and
$\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
properties:
\begin{enumerate}
\item Closure under simultaneous substitutions: if $\rho \in
  \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow \Delta$,
  then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
\item If we have a pair of relation environments $\rho_1 \in
  \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
  \relEnv{E}(\Delta)$, along with a simultaneous substitution $\Delta'
  \vdash \sigma \Rightarrow \Delta$, such that $\rho_1 \circ
  \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists a $\rho
  \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho \circ
  \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ \pi_{i\mathord:s'} =
  \rho_2$.
  % such that the outer edges of the
  % following diagram commute, for all primitive types $\tyPrimNm{X} \in
  % \mathit{PrimType}$:
  % %FIXME: check that these compositions are the right way round
  % %FIXME: re-do this as a set of equations
  % %FIXME: give a better English description of why this is needed
  % \begin{displaymath}
  %   \xymatrix{
  %     {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
  %     &
  %     {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
  %     \\
  %     {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
  %     &
  %     {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
  %     \\
  %     &
  %     &
  %     {\Rel(\tyPrimSem{\tyPrimNm{X}})}
  %   }
  % \end{displaymath}
  % Then there exists a relation environment $\rho \in
  % \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
  % two triangles in the bottom right of the diagram commute.
  % FIXME: mention that we already know that the square commutes
\end{enumerate}

@@

% \begin{example}
%   The construction of relational environments is the most subtle part
%   of our development. \fixme{complete this part}
% \end{example}

% \begin{example}[Units of Measure]
%   We give two examples of assignments of sets of relational
%   environments to index contexts in the units of measure example. The
%   second example is a refinment of the first. FIXME: some general
%   words about relational environments in these cases.
%   \begin{enumerate}
%   \item FIXME: assign a positive scaling factor to each variable in an
%     index context, extend homomorphically to all index expressions,
%     and then use this to define a scaling relation
%   \item FIXME: given an index context, assume a subgroup and a
%     homomorphism into the scaling factors
%   \end{enumerate}
% \end{example}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Well-typed Programs}
\label{sec:well-typed-programs}

>>>>

\begin{example}[Two-Dimensional Geometry]
  \label{ex:two-dim-geo-operations}
  The two-dimensional geometry system has a sort for each of the
  geometric groups that we mentioned in
  \autoref{sec:motivating-examples}. Thus, $\SortSet =
  \{\SynTransl{2}, \SynGL{2}, \SynOrth{2}, \SynGL{1} \}$. For the
  index operations, we have additive group structure on
  $\SynTransl{2}$, multiplicative group structure on $\SynGL{1}$,
  $\SynGL{2}$ and $\SynOrth{2}$, injections from $\SynOrth{2}$ and
  $\SynGL{1}$ into $\SynGL{2}$, determinant, and absolute value. Thus,
  $\IndexOpSet = \{ 0, +, -, 1_G, -\cdot_G-, -^{-1_G}, \iota_O,
  \iota_1, \det, |\cdot| \}$, where $G \in \{\SynGL{1}, \SynGL{2},
  \SynOrth{2}\}$, with the following assignment of arities:
  \begin{displaymath}
    \begin{array}{@{}l@{\hspace{0em}=\hspace{0em}}l@{\hspace{0.5em}}l@{\hspace{0em}=\hspace{0em}}l}
      \indexOpArity(0) & ([], \SynTransl{2}) &
      \indexOpArity(1_G) & ([], G) \\
      \indexOpArity(+) & ([\SynTransl{2}, \SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(\cdot_G) & ([G,G],G) \\
      \indexOpArity(-) & ([\SynTransl{2}], \SynTransl{2}) &
      \indexOpArity(^{-1_G}) & ([G], G) \\
      \indexOpArity(\iota_o) & ([\SynOrth{2}], \SynGL{2}) &
      \indexOpArity(\iota_1) & ([\SynGL{1}], \SynGL{2}) \\
      \indexOpArity(\det) & ([\SynGL{2}], \SynGL{1}) &
      \indexOpArity(|\cdot|) & ([\SynGL{1}], \SynGL{1})
    \end{array}
  \end{displaymath}
  The intended interpretations of the top three pairs of operations
  are group unit, group combination and group negation,
  respectively. When we introduce equational theories for index
  expressions in \autoref{sec:type-equality} below, we will impose the
  (abelian) group laws. For this example, we have $\PrimTypeSet = \{
  \tyPrimNm{vec}, \tyPrimNm{real} \}$, with
  $\primTyArity(\tyPrimNm{vec}) = [\SynGL{2}, \SynTransl{2}]$ and
  $\primTyArity(\tyPrimNm{real}) = [\SynGL{1}]$.
\end{example}

@@

  {\small
  \textbf{Well-sorted index expressions}
  \begin{mathpar}
    \inferrule* [right=IVar]
    {i : s \in \Delta}
    {\Delta \vdash i : s}

@@

    \inferrule* [right=TyEx]
    {\Delta, i \mathord: s \vdash A \isType}
    {\Delta \vdash \exists i \mathord: s. A \isType}
  \end{mathpar}}
  \caption{Index expressions and types}
  \label{fig:indexes-and-types}
\end{figure*}

Given the above data, the rules in \autoref{fig:indexes-and-types}
generate two judgements: well-sorted index expressions $\Delta \vdash
e : s$ and well-indexed types $\Delta \vdash A \isType$. Since index
variables may appear in types, types are judged to be well-indexed
with respect to an index context $\Delta$. The rules for well-sorted
index expressions are particularly simple: either an index expression
is a variable that appears in the context (rule \TirName{IVar}), or it
is an application of an index operation taken from $\IndexOpSet$ to
other index expressions (rule \TirName{IOp}). The rules for
well-indexed types include the usual rules for constructing types of
the simply-typed $\lambda$-calculus with unit, sum and tuple types
(rules \TirName{TyUnit}, \TirName{TyArr}, \TirName{TyTuple} and
\TirName{TySum}). The rule \TirName{TyPrim} allows us to form, from a
a primitive type $\tyPrimNm{X}$ and appropriately sorted index
expressions $e_1,...,e_n$, the well-indexed type
$\tyPrim{X}{e_1,...,e_n}$. The rule \TirName{TyForall} permits the
formation of universally quantified types, where the universal
quantification ranges over all index expressions of some
sort. Existential types are constructed using the \TirName{TyEx}
rule, and allow for abstraction by hiding. %FIXME: think about the introduction of existentials more, perhaps forward ref to their use

@@

\begin{example}[Two-Dimensional Geometry]
  \label{ex:two-dim-geo-axioms}
  In \autoref{sec:motivating-examples}, we assumed that various
  equational axioms hold for the indexing expressions standing for
  elements of geometric groups. We can now make this assumption formal
  in our general framework. Semantically, translations form an abelian
  group under addition, so we assume the abelian group axioms for
  translations:
  \begin{displaymath}
    \begin{array}{l}
      t : \SynTransl{2} \vdash t + 1 \stackrel{ax}\equiv t : \SynTransl{2} \\
      t_1, t_2, t_3 : \SynTransl{2} \vdash t_1 + (t_2 + t_3) \stackrel{ax}\equiv (t_1 + t_2) + t_3 : \SynTransl{2} \\
      t : \SynTransl{2} \vdash t + (-t) \stackrel{ax}\equiv 1 : \SynTransl{2} \\
      t_1, t_2 : \SynTransl{2} \vdash t_1 + t_2 \stackrel{ax}\equiv t_2 + t_1 : \SynTransl{2} \\
    \end{array}
  \end{displaymath}
  Similarly, the sort of scale factors $\SynGL{1}$ forms an abelian
  group under multiplication, and the sorts $\SynGL{2}$ and
  $\SynOrth{2}$ form (non-abelian) multiplicative groups, so we assume
  the appropriate axioms. We also assume that the operations $\iota_O,
  \iota_1, \det$ and $|\cdot|$ are group homomorphisms, and that
  expressions of the form $\iota_1(s)$ commute with group
  multiplication in the sort $\SynGL{2}$. The absolute value of the
  determinant of an orthgonal transformation is always $1$, so we also
  assume $|\det(\iota_OO)| \stackrel{ax}\equiv 1$. Likewise, scaling
  maps have a determined determinant: $\det(\iota_1(s))
  \stackrel{ax}\equiv s \cdot s$.
\end{example}

@@

The defining feature of the index erasure interpretation is that the
semantics of a well-indexed type $\tyPrim{X}{e_1,...,e_n}$ is
determined solely by the primitive type $\tyPrimNm{X}$ and not by the
index expressions $e_1,...,e_n$. Accordingly, we assume that each
primitive type $\tyPrimNm{X} \in \PrimTypeSet$ is assigned a set
$\tyPrimSem{\tyPrimNm{X}}$. We extend this assignment to every
well-indexed type by induction on the type structure:
\begin{displaymath}
  \begin{array}{@{\hspace{0em}}c@{\hspace{0em}}c}
    \begin{array}{@{}l}
      \tySem{\tyUnit} \isDefinedAs \{*\} \\
      \tySem{A \tyProduct B} \isDefinedAs \tySem{A} \times \tySem{B} \\
      \tySem{A \tyArr B} \isDefinedAs \tySem{A} \to \tySem{B} \\
    \end{array}
    &
    \begin{array}{l}
      \tySem{A + B} \isDefinedAs \tySem{A} + \tySem{B} \\
      \tySem{\tyPrim{X}{e_1,...,e_n}} \isDefinedAs \tyPrimSem{\tyPrimNm{X}} \\
      \tySem{\forall i\mathord:s. A} \isDefinedAs \tySem{A} \\
      \tySem{\exists i\mathord:s. A} \isDefinedAs \tySem{A}
    \end{array}
  \end{array}
\end{displaymath}
% The interpretation of the unit type is a chosen one element set, and
% the interpretations of the function, tuple and sum types are simply
% the corresponding constructions on sets. 
The interpretations of the
universal and existential quantifiers ignore the indexing: the
interpretation of the type $\forall i\mathord:s.A$ is exactly the
interpretation of the type $A$, and likewise for $\exists
i\mathord:s.A$.

@@

@@

For the unit, tuple, sum and function types we will define the
relational interpretation as a standard logical relation. The
relational interpretations of primitive types with index arguments and
the universally quantified types require an interpretation of index
contexts. We assign interpretations to index contexts in terms of sets
of \emph{relational environments}.

The definition of relational environments is the most subtle part of
our development. The straightforward approach would be to assume some
interpretation of the indexing expressions, where sorts are
interpreted as sets and index operations as functions, and then define
the relational interpretation of a primtive type
$\tyPrim{X}{e_1,...,e_n}$ in terms of the interpretations of
$e_1,...,e_n$. This approach suffices to prove free theorems and type
isomorphisms for instantiations of our general framework. However, in
order to prove non-definability results (like, for example, the one in
\exref{ex:uninhabited-type}), we choose a more subtle interpretation
of index contexts and primitive types that can account more precisely
for the definable programs permitted by the algebraic indexing. In
this section, we set out the general requirements that we require of
relational environments in order for the abstraction theorem to
hold. In \autoref{sec:constr-rel-env}, we give a general construction
of relational environments that will allow us to prove several
non-definability results.

A relational environment for a context $\Delta$ is a (dependently typed)
function $\rho$ that, for each primitive type $\tyPrimNm{X}$ and
instantiation of its arguments $\sigma$, assigns a binary relation
$\rho\ \tyPrimNm{X}\ \sigma$ on the index-erasure interpretation of
$\tyPrimNm{X}$:
\begin{displaymath}
  \rho : (\tyPrimNm{X} \in \PrimTypeSet) \to (\Delta \Rightarrow \primTyArity(\tyPrimNm{X})) \to \Rel(\tyPrimSem{\tyPrimNm{X}})
\end{displaymath}
Every relation environment must respect index expression equality: for
any $\tyPrimNm{X} \in \PrimTypeSet$ and pair of simultaneous
substitutions $\Delta \vdash \sigma \Rightarrow
\primTyArity(\tyPrimNm{X})$ and $\Delta \vdash \sigma' \Rightarrow
\primTyArity(\tyPrimNm{X})$, if $\Delta \vdash \sigma \equiv \sigma'
\Rightarrow \primTyArity(\tyPrimNm{X})$, then $\rho\ \tyPrimNm{X}\
\sigma = \rho\ \tyPrimNm{X}\ \sigma'$.

@@

As we stated above, index contexts $\Delta$ are interpreted as certain
sets of relation environments, i.e.,~subsets of
$\mathrm{RelEnv}(\Delta)$. Depending on the primitive operations that
we assume for our system, we may have different sets of relational
environments that enforce different invariants. We require that the
sets of relational environments are \emph{valid} in the following
sense:

\begin{definition}
  \label{defn:valid-rel-env-family}
  A family of sets of relational environments $\relEnv{E}(\Delta)
  \subseteq \mathrm{RelEnv}(\Delta)$ is \emph{valid} if the following
  two properties are satisfied:
  \begin{enumerate}
  \item Closure under simultaneous substitutions: if $\rho \in
    \relEnv{E}(\Delta)$ and $\Delta' \vdash \sigma \Rightarrow \Delta$,
    then $\rho \circ \sigma \in \relEnv{E}(\Delta')$.
  \item If we have a pair of relation environments $\rho_1 \in
    \relEnv{E}(\Delta', i\mathord:s')$ and $\rho_2 \in
    \relEnv{E}(\Delta)$, along with a simultaneous substitution
    $\Delta' \vdash \sigma \Rightarrow \Delta$, such that $\rho_1
    \circ \pi_{i\mathord:s'} = \rho_2 \circ \sigma$, then there exists
    a $\rho \in \relEnv{E}(\Delta, i\mathord:s')$ such that $\rho
    \circ \sigma_{i\mathord:s'} = \rho_1$ and $\rho \circ
    \pi_{i\mathord:s'} = \rho_2$.
    % such that the outer edges of the
    % following diagram commute, for all primitive types $\tyPrimNm{X} \in
    % \mathit{PrimType}$:
    % % FIXME: check that these compositions are the right way round
    % % FIXME: re-do this as a set of equations
    % % FIXME: give a better English description of why this is needed
    % \begin{displaymath}
    %   \xymatrix{
    %   {\Delta \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar[d]_{- \circ \sigma}
    %   &
    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[d]^{- \circ \sigma_{s'}} \ar@/^/[rdd]^{\rho_1 \tyPrimNm{X}}
    %   \\
    %   {\Delta' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar[r]^(.45){- \circ \pi_{i\mathord:s'}} \ar@/_/[rrd]_{\rho_2\ \tyPrimNm{X}}
    %   &
    %   {\Delta,i\mathord:s' \Rightarrow \primTyArity(\tyPrimNm{X})} \ar@{.>}[dr]_{\rho\ \tyPrimNm{X}}
    %   \\
    %   &
    %   &
    %   {\Rel(\tyPrimSem{\tyPrimNm{X}})}
    % }
    % \end{displaymath}
    % Then there exists a relation environment $\rho \in
    % \relEnv{E}(\Delta,i\mathord:s')$ (the dotted arrow) such that the
    % two triangles in the bottom right of the diagram commute.
    % FIXME: mention that we already know that the square commutes
  \end{enumerate}
\end{definition}

% For the general
% framework, we assume that, for each index context $\Delta$, we are
% given a set of relation environments $\relEnv{E}(\Delta)$. Note that
% this assignment of sets of relation environments to index contexts is
% not assumed to be compositional: it is not necessarily the case that
% the set of relational environments $\relEnv{E}(\Delta_1,\Delta_2)$ is
% defined in terms of $\relEnv{E}(\Delta_1)$ and
% $\relEnv{E}(\Delta_2)$. However, it must satisfy the following two
% properties:

@@

@@

\subsubsection{A General Construction of Relational Environments}
\label{sec:constr-rel-env}

\newcommand{\Free}{\mathrm{Free}}
\newcommand{\semSort}[1]{\llbracket #1 \rrbracket^{\mathcal{S}}}
\newcommand{\semIndexExp}[1]{\llbracket #1 \rrbracket^{\mathcal{I}}}

We now give a general construction of valid families of sets of
relational environments that will allow us to derive useful
consequences of the abstraction theorem, including free theorem, type
isomorphisms and non-definability results. We make use of some
concepts of universal algebra, which we recall here. \fixme{Desperately needs some intuition: do so in terms of clumpiness}

\paragraph{Some Universal Algebra}
For each sort $s \in \SortSet$ we assume a carrier set
$\semSort{s}$. We extend this assignment to index contexts and other
sequences of sorts (e.g., type arities) using the cartesian product of
sets: $\semSort{i_1\mathord:s_1,...,i_n\mathord:s_n} = \semSort{s_1}
\times ... \times \semSort{s_n}$. For each index operation $\texttt{f}
\in \IndexOpSet$ with $\indexOpArity(\texttt{f}) = ([s_1,...,s_n],s)$,
we assume a function $\semIndexExp{\texttt{f}} : \semSort{s_1,...,s_n}
\to \semSort{s}$. For each well-sorted index expression $\Delta \vdash
e : s$, we assign a function $\semIndexExp{e} : \semSort{\Delta} \to
\semSort{s}$ by recursion on the structure of $e$. We further assume
that for each axiom $\Delta \vdash e \stackrel{ax}\equiv e' : s \in
\IndexAxiomSet$, we have $\semIndexExp{e} = \semIndexExp{e'}$. A
\emph{model} is a pair $\mathcal{M} = (\semSort{\cdot}_{\mathcal{M}},
\semIndexExp{\cdot}_{\mathcal{M}})$ such that all the axioms in
$\IndexAxiomSet$ are satisfied.

% A sub-model $\mathcal{N}$ of a given
% model $\mathcal{M}$ (written $\mathcal{N} \subseteq \mathcal{M}$) is a
% model such that $\semSort{s}_{\mathcal{N}} \subseteq
% \semSort{s}_{\mathcal{M}}$ for all $s \in \SortSet$.

\fixme{Define free models for any sort-indexed collection of sets}
Given an index context $\Delta$, the \emph{free} model on $\Delta$,
$\Free(\Delta)$, has $\semSort{s}_{\Free(\Delta)} = \{e \sepbar \Delta
\vdash e : s\}/\mathord\equiv$: the set of expressions of sort $s$ in
the indexing context $\Delta$, quotiented by the equality
judgment. The free model has
$\semIndexExp{\texttt{f}}_{\Free(\Delta)}(e_1,..,e_n) =
\texttt{f}(e_1,...,e_n)$.

Given a pair of models $\mathcal{M}_1 = (\semSort{\cdot}_1,
\semIndexExp{\cdot}_1)$ and $\mathcal{M}_2 = (\semSort{\cdot}_2,
\semIndexExp{\cdot}_2)$, a \emph{homomorphism} $h : \mathcal{M}_1 \to
\mathcal{M}_2$ is a family of functions $\{ h_s : \semSort{s}_1 \to
\semSort{s}_2 \}_{s \in \SortSet}$ such that for each $\texttt{f} \in
\IndexOpSet$, we have $h_s(\semIndexExp{\texttt{f}}_1 (x_1,...,x_n)) =
\semIndexExp{\texttt{f}}_2(h_{s_1}x_1, ..., h_{s_n}x_n)$.

\paragraph{Relational Environments} \fixme{Say some stuff in english to make sure that the construction is understandable}

To assign a relational meaning to
each of the primitive types we assume a some model $\mathcal{M}$ of
the indexing theory along with a function $R_{\tyPrimNm{X}} :
\semSort{\primTyArity(\tyPrimNm{X})}_{\mathcal{M}} \to
\Rel(\tyPrimSem{\tyPrimNm{X}})$ for each primitive type
$\tyPrimNm{X}$, and a constant $R^\bullet_{\tyPrimNm{X}} \in
\Rel(\tyPrimSem{\tyPrimNm{X}})$. We set
$\relEnv{E}_{\mathcal{M},R}(\Delta) = \{ \rho^{\mathcal{A},h} \sepbar \mathcal{A}
\subseteq \Free(\Delta), h : \mathcal{A} \to \mathcal{M} \}$, where:
\begin{displaymath}
  \rho^{\mathcal{A},h}\ \tyPrimNm{X}\ (e_1,...,e_n) = \left\{
    \begin{array}{ll}
      R^\bullet_{\tyPrimNm{X}} & \textrm{if any }e_i \not\in \mathcal{A} \\
      R_{\tyPrimNm{X}}(h(e_1),...,h(e_n)) & \textrm{if all }e_i \in \mathcal{A}
    \end{array}
  \right.
\end{displaymath}

\begin{lemma}
  The family $\relEnv{E}_{\mathcal{M},R}$ of sets of relation
  environments is valid (c.f.~\defref{defn:valid-rel-env-family}).
\end{lemma}

% \fixme{Note that we can recover the basic approach by restricting the
%   sub-algebras to just the whole thing. Does this work in general?
%   What about arbitrary collections of sub-algebras?}
